---
title: "glm+gam practical"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## glm & gam practical

### poisson glm

*1. Import sorich.csv*

```{r task one}
library(tidyverse)

sorich <- read_csv('sorich.csv')
view(sorich)

ggplot(sorich, aes(x = cover, y = nsp)) + geom_point()

```

*2. Fit a linear model for species richness against cover assuming a Normal distribution. *

``` {r task two}

fit_lm <- glm(nsp ~ cover, data = sorich)
anova(fit_lm)
summary(fit_lm)

```

P-value is < 2e-16

``` {r task two}

ggplot(cbind(sorich, fit = fitted(fit_lm)), aes(x = cover, y = nsp)) + 
  geom_point() + 
  geom_line(aes(y = fit), colour = "red") +
  labs(y = "Species richness", x = "Cover")

```

*3. Study the diagnostic plots*

For the linear model:

``` {r task three}

par(mfrow=c(2,2))
plot(fit_lm)

```

Interpreting the diagnostic plots:
Residuals vs Fitted: almost a straight line, good!
Normal Q-Q: not following the line at the top, not good. I think this means that the residuals are not normally distributed. 
Scale-Location: almost a straight line, good!
Residuals vs Leverage: all points are within the dotted lines, good!

For the glm model:



*4. What distribution is the response?*

The distribution for the response is Poisson because there is a categorical predictor. 

*5. Do an appropriate analysis*

Since the distribution of the response is not normal, but Poisson, we use a generalised linear model. 

``` {r task five}

fit_glm <- glm(nsp ~ cover, data = sorich, family = poisson)
anova(fit_glm, test = "Chisq")
summary(fit_glm)

```

The p-value in the summary is < 2e-16. 

``` {r task two}

ggplot(cbind(sorich, fit = fitted(fit_glm)), aes(x = cover, y = nsp)) + 
  geom_point() + 
  geom_line(aes(y = fit), colour = "blue") +
  labs(y = "Species richness", x = "Cover")

```

*6. Check for over-dispersion*

``` {r task six}

anova(fit_glm)

```

*7. Interpret the results*

Since the residual deviance (1315) is (much) larger than the residual df (690), there is overdispersion. 

*8. How does the width of the confidence interval at cover = 10 change when over-dispersion is allowed for*

``` {r task eight}

pred <- predict(fit_glm, newdata = data.frame(cover = 10), type = "confidence interval")
view(pred)

```



9. Do the grasses differ from the herb, i.e. include the factor grasherb and test its significance?