---
title: "test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an *R Markdown document*. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

<!-- Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. -->

#This is a header
## this is another header
**bold** *italics*
`r 27 * 84`
$A = \pi*r^{2}$

``` {r iris, fig.cap="table showing iris data"}
library(tidyverse)
as_tibble(iris)
head(iris, 3) %>% knitr::kable(caption = "A kable table")
```


```{r cars-figure, fig.cap="Plot showing relationship ..."}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
```

this is a reference to the car figure \@ref(fig:cars-figure))